Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2024-12-29 src/docs_paper-jtb/appendix.Rmd
Format Section 5.4.1
Label equations by default and some edits

The proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ was modelled as
\begin{equation}
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
\end{equation}
where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
The prior distribution for $\Omega_T$ was informed by the characteristics of the recent infection testing algorithm.
For PHIA surveys this was $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
For PHIA surveys there was assumed to be no false recency, such that $\beta_{T_0} = 0.0$, $\sigma^{\beta_T} = 0.0$, and $\beta_T = 0$.

@@

(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A). As a result, the number of iterations required for the the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).

@@

(ref:sd-alt-output) Differences in Naomi model output posterior standard deviations as estimated by GEB and GPCA-AGHQ compared to NUTS. Each point is an estimate of the indicator for a particular strata. Error is increased by GPCA-AGHQ for HIV prevalence and HIV incidence, and reduced for ART coverage.

>>>>

The proportion who were recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ was modelled as
\begin{equation}
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
\end{equation}
where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
The prior distribution for $\Omega_T$ was informed by the characteristics of the recent infection testing algorithm.
For PHIA surveys this was $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
For PHIA surveys there was assumed to be no false recency, such that $\beta_{T_0} = 0.0$, $\sigma^{\beta_T} = 0.0$, and $\beta_T = 0$.

@@

(ref:ess) The efficiency of the NUTS, as measured by the ratio of effective sample size to total number of iterations run, was low for most parameters (Panel A). As a result, the number of iterations required for the effective number of samples (mean `r round(naomi_nuts$ess_mean)`) to be satisfactory was high (Panel B).

@@

(ref:sd-alt-output) Differences in Naomi model output posterior standard deviations as estimated by GEB and GPCA-AGHQ compared to NUTS. Each point is an estimate of the indicator for a particular strata. Error is increased by GPCA-AGHQ for ART coverage and HIV incidence, and reduced for HIV prevalence.

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Label equations by default and some edits

---
title: Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    correspondingauthor: true
  - name: Alex Stringer
    email: alex.stringer@uwaterloo.ca
    affiliation: B
  - name: Seth R. Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    affiliation: C
  - name: Jeffrey W. Imai-Eaton
    email: jeaton@hsph.harvard.edu
    affiliation: D
address:
  - code: A
    organization: Department of Mathematics
    country: Imperial College London
  - code: B
    organization: Department of Statistics and Actuarial Science
    country: University of Waterloo
  - code: C
    organization: Department of Computer Science
    country: University of Oxford
  - code: D
    organization: Harvard T.H. Chan School of Public Health
    country: Harvard University
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. The model is provided as a tool for countries to input their data to and generate estimates with during a yearly process supported by UNAIDS. Previously, inference has been conducted using empirical Bayes and a Gaussian approximation, implemented via the \texttt{TMB} \textsc{R} package. We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with more than 20 hyperparameters. Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler. Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
keywords: 
  - Bayesian statistics
  - spatial statistics
  - evidence synthesis
  - small-area estimation
  - approximate inference
  - INLA
  - AGHQ
  - HIV epidemiology
journal: "Journal of Theoretical Biology"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: citations.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    number_sections: yes
    keep_tex: true
    citation_package: natbib
    includes:
      in_header: preamble.tex
---

@@

The Naomi small-area estimation model [@eaton2021naomi; @esra2024improved] addresses these challenges by synthesising data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
Modelling multiple data sources jointly mitigates the limitations of any single source and increases statistical power.
Software developed for Naomi (\url{https://naomi.unaids.org}) facilitates over 35 countries inputting their data and interactively generating estimates during workshops as a part of a yearly estimates process supported by UNAIDS.
Generation of estimates by country teams is an important and distinctive feature of the HIV response.
Drawing on expert knowledge of the data improves the accuracy of the process and strengthens trust in the resulting estimates, creating a virtuous cycle of data quality, use, and ownership [@noor2022country].

Naomi comprises multiple linked generalized linear mixed models (GLMMs) and presents a challenging Bayesian inference problem.
The model contains hundreds of fixed and random effect parameters and over 20 hyperparameters.
This is substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
Additionally, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
Inference must be fast and have low memory usage, to allow interactive review and iteration of model results by workshop participants.
The scale of the model and features of its posterior geometry [@neal2003slice] make Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
Inference must be reliable and automatic across a range of country settings, and should not require close monitoring and diagnostics (e.g. for MCMC convergence).

@@

Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $[n] = \{1, \ldots, n\}$.
The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.

@@

Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of non-linearity in the model.
Put together, ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
\end{align*}
with latent field and hyperparameter priors as in the LGM case.

@@

These processes are each informed by household survey data.
Weighted aggregate survey observations are calculated based on individual responses $\theta_j \in \{0, 1\}$ as
\begin{equation*}
\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}.
\end{equation*}
Design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$ are supplied by the survey provider and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
The weighted observed number of outcomes is taken to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
\begin{equation*}
m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
\end{equation*}
is the Kish effective sample size [ESS; @kish1965survey].
The weighted aggregated number of outcomes are modelled using a binomial working likelihood [@chen2014use] defined to operate on the reals
\begin{equation*}
y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
\end{equation*}
The terms $\theta_{I}$ are the following weighted aggregates
\begin{equation*}
\rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
\alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
\end{equation*}

@@

Women attending ANC clinics are routinely tested for HIV, to help prevent mother-to-child transmission.
HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
\end{align*}
These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
The number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
The number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ are modelled using nested binomial likelihoods
\begin{align*}
y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
\end{align*}
As in the household survey component, weighted aggregates
\begin{equation*}
\rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
\alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
\end{equation*}
are used, with $\Psi_i$ the number of pregnant women assumed here to be fixed.

@@

The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ and hyperparameter prior distribution $p(\btheta)$ in an ELGM is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
\end{align}

@@

AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1), and conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
AS was supported by NSERC Discovery Grant RGPIN-2023-03331.
SRF was supported by the EPSRC (EP/V002910/2).
JWE was supported by UNAIDS and National Institute of Allergy and Infectious Disease of the National Institutes of Health (R01AI136664).
This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.

>>>>

---
title: Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    correspondingauthor: true
  - name: Alex Stringer
    email: alex.stringer@uwaterloo.ca
    affiliation: B
  - name: Seth R. Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    affiliation: C
  - name: Jeffrey W. Imai-Eaton
    email: jeaton@hsph.harvard.edu
    affiliation: D
address:
  - code: A
    organization: Department of Mathematics
    country: Imperial College London
  - code: B
    organization: Department of Statistics and Actuarial Science
    country: University of Waterloo
  - code: C
    organization: Department of Computer Science
    country: University of Oxford
  - code: D
    organization: Harvard T.H. Chan School of Public Health
    country: Harvard University
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. The model is provided as a tool for countries to input their data to and generate estimates with during a yearly process supported by UNAIDS. Previously, inference has been conducted using empirical Bayes and a Gaussian approximation, implemented via the \texttt{TMB} \textsc{R} package. We propose a new inference method based on an extension of adaptive Gauss-Hermite quadrature to deal with more than 20 hyperparameters. Using data from Malawi, our method improves the accuracy of inferences for model parameters, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler. Our implementation leverages the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
keywords: 
  - Bayesian statistics
  - spatial statistics
  - evidence synthesis
  - small-area estimation
  - approximate inference
  - INLA
  - AGHQ
  - HIV epidemiology
journal: "Journal of Theoretical Biology"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: citations.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    number_sections: yes
    keep_tex: true
    citation_package: natbib
    includes:
      in_header: preamble.tex
---

@@

The Naomi small-area estimation model [@eaton2021naomi; @esra2024improved] addresses these challenges by synthesising data from multiple sources to estimate HIV indicators at a district-level, by age and sex.
Modelling multiple data sources jointly mitigates the limitations of any single source and increases statistical power.
Software developed for Naomi (\url{https://naomi.unaids.org}) facilitates over 35 countries inputting their data and interactively generating estimates during workshops as a part of a yearly estimates process supported by UNAIDS.
Generation of estimates by country teams is an important and distinctive feature of the HIV response.
Drawing on expert knowledge about the data improves the accuracy of the process and strengthens trust in the resulting estimates, creating a virtuous cycle of data quality, use, and ownership [@noor2022country].

Naomi comprises multiple linked generalized linear mixed models (GLMMs) and presents a challenging Bayesian inference problem.
The model contains hundreds of fixed and random effect parameters and over 20 hyperparameters.
This is substantially more than the small number that can typically be handled by approaches like integrated nested Laplace approximations [INLA; @rue2009approximate].
Additionally, observations depend on multiple structured additive predictors, such that Naomi falls into the class of extended latent Gaussian models [ELGMs; @stringer2022fast].
Inference must be fast and have low memory usage, to allow interactive review and iteration of model results by workshop participants.
The scale of the model and features of its posterior geometry [@neal2003slice] make Markov chain Monte Carlo (MCMC) approaches are prohibitively slow.
Inference should also be reliable, automatic across various country settings, and require minimal manual monitoring (e.g. of MCMC convergence).

@@

Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with
\begin{align}
y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align}
where $[n] = \{1, \ldots, n\}$.
The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.

@@

Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of non-linearity in the model.
Put together, ELGMs are then of the form
\begin{align}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
\end{align}
with latent field and hyperparameter priors as in the LGM case.

@@

These processes are each informed by household survey data.
Weighted aggregate survey observations are calculated based on individual responses $\theta_j \in \{0, 1\}$ as
\begin{equation}
\hat \theta_I = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}.
\end{equation}
Design weights $w_j$ for each of $\theta \in \{\rho, \alpha, \kappa\}$ are supplied by the survey provider and aim to reduce bias by decreasing possible correlation between response and recording mechanism [@meng2018statistical].
The index $j$ runs across all individuals in strata $i \in I$ within the relevant denominator i.e. for ART coverage, only those individuals who are HIV positive.
The weighted observed number of outcomes is taken to be $y^{\theta}_{I} = m^{\theta}_{I} \cdot \hat \theta_{I}$ where
\begin{equation}
m^{\theta}_I = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
\end{equation}
is the Kish effective sample size [ESS; @kish1965survey].
The weighted aggregated number of outcomes are modelled using a binomial working likelihood [@chen2014use] defined to operate on the reals
\begin{equation}
y^{\theta}_{I} \sim \text{xBin}(m^{\theta}_{I}, \theta_{I})
\end{equation}
The terms $\theta_{I}$ are the following weighted aggregates
\begin{equation}
\rho_{I} = \frac{\sum_{i \in I} N_i \rho_i}{\sum_{i \in I} N_i}, \quad
\alpha_{I} = \frac{\sum_{i \in I} N_i \rho_i \alpha_i}{\sum_{i \in I} N_i \rho_i}, \quad
\kappa_{I} = \frac{\sum_{i \in I} N_i \rho_i \kappa_i}{\sum_{i \in I} N_i \rho_i}.
\end{equation}

@@

Women attending ANC clinics are routinely tested for HIV, to help prevent mother-to-child transmission.
HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
\begin{align}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
\end{align}
These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
The number of ANC clients with ascertained status to be fixed as $m^{\rho^\text{ANC}}_I$. 
The number of those with positive status $y^{\rho^\text{ANC}}_I$, and the number of those already on ART prior to their first ANC visit $y^{\alpha^\text{ANC}}_I$ are modelled using nested binomial likelihoods
\begin{align}
y^{\rho^\text{ANC}}_I &\sim \text{Bin}(m^{\rho^\text{ANC}}_I, \rho^\text{ANC}_{I}), \\
y^{\alpha^\text{ANC}}_I &\sim \text{Bin}(y^{\rho^\text{ANC}}_I, \alpha^\text{ANC}_{I}).
\end{align}
As in the household survey component, weighted aggregates
\begin{equation}
\rho^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in I} \Psi_i}, \quad
\alpha^\text{ANC}_{I} = \frac{\sum_{i \in I} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in I} \Psi_i \rho_i^\text{ANC}},
\end{equation}
are used, with $\Psi_i$ the number of pregnant women assumed here to be fixed.

@@

The joint posterior of the parameters $(\x, \btheta)$ given data $\y$ and hyperparameter prior distribution $p(\btheta)$ in an ELGM is given by
\begin{equation}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, \x_{\mathcal{J}_i}, \btheta) \right).
\end{equation}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &\approx \tilde p(x_i \, | \, \y) = \int \tilde p(x_i \, | \, \btheta, \y) \tilde p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &\approx \tilde p(\theta_j \, | \, \y) = \int \tilde  p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
\end{align}

@@

AH was supported by the EPSRC Centre for Doctoral Training in Modern Statistics and Statistical Machine Learning (EP/S023151/1).
AH conducted part of this research while an International Visiting Graduate Student at the University of Waterloo.
AH and JWE were supported by the Bill and Melinda Gates Foundation (OPP1190661, OPP1164897).
AS was supported by NSERC Discovery Grant (RGPIN-2023-03331).
SRF was supported by the EPSRC (EP/V002910/2).
JWE was supported by UNAIDS and National Institute of Allergy and Infectious Disease of the National Institutes of Health (R01AI136664).
This research was supported by the MRC Centre for Global Infectious Disease Analysis (MR/R015600/1), jointly funded by the UK Medical Research Council (MRC) and the UK Foreign, Commonwealth \& Development Office (FCDO), under the MRC/FCDO Concordat program and is also part of the EDCTP2 programme supported by the European Union.
